Pythran: Bringing OpenMP to a Python Subset

### 2013-02-06 doc/papers/europar2013/paper.tex
Major Paper update.
Take into account A. Cohen's review.

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}

\lstset{language=python}

@@

    The Python language offers an interesting alternative as a scientific
    computing language thanks to the SciPy package. However, the lack of
    fine-grained parallelism support remains an important shortcoming. This
    paper proposes to add OpenMP directives to the Python language and shows
    that it is possible to turn a module written in a large Python subset
    enhanced with these directive into a native parallel module that runs as
    fast as its naive C equivalent. The approach is validated through the
    Pythran compiler.

@@

The prohibitive performance overhead of the language implied by its interpreted
and highly dynamic nature should prevent its usage as a language for high
performance computing. SciPy overcomes this issue through the usage of low-level
routines written in C or Fortran and encapsulated in Python through
\emph{native} modules. Moreover, its core data structure, the multi-dimensional
array~\cite{numpyarray2011}, was designed so that the underlying data are
available to both native modules and the Python interpreter without conversion
costs.

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
the existing approaches and shows a critical lack of parallelism support.
Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives
to the Python language in order to add fine-grain parallelism support to Python
in the context of a Python subset called Pythran. Static compilation of this
language into C++ and the underlying runtime is described in
Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
scientific applications and compared to both interpreted and native code in
Section~\ref{sec:validation}.

@@

In the many cores area, turning to parallelism to balance the performance
limitations of scripting languages, as described in~\cite{choy05} in the context
of the MATLAB language, or in~\cite{mals07} in the context of the R language,
seems legitimate. In the context of Python, most approaches have
focused on fork-based parallelism.

@@

As a consequence, Python developers need to write multi-threaded native
modules in order to fully benefit from multiple cores. This lead to a kind of
computations referred as hybrid computations.

@@

In the context of interpreted language, an hybrid computation is defined
in~\cite{dongara2007} as a computation where part of the code is interpreted,
and part of the code is executed natively.

It is now common for scripting language to have C bindings. To take advantage of
compiled code, and to overcome the GIL limitations, Python developers must write
parallel C/C++ functions and the associated boilerplate based on the Python C
API~\cite{pythoncapi}. Tools have been developed to relieve the user from
writing the encapsulating code, notably Swig~\cite{swig2003} that relies on an
enhanced interface specification, or
\texttt{boost::python}~\cite{boostpython2007} that relies on C++ facilities to
guide translation.

@@

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, among which the possibility to define
parallel loops, but no task parallelism. Plw~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.  The
numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
uses additional type information to generate sequential LLVM bytecode.  PyCUDA
and PyOpenCL~\cite{klockner2012} also mixes python with kernels embeded as raw
strings to target accelerators.

The core issue of these approaches is that they imply to modify, in a more or
less trivial way, the original code. They require to invest in a new language,
and the long-term preservation of this investment is not ensured.

@@

Copperhead~\cite{shedskin2006} is a functional, data parallel language embedded in
Python. It uses n-uplets, Numpy arrays and lists as its core data structure and
prohibits usage of many control-flow operators such as loops, enforcing the use
of the \texttt{map}, \texttt{filter} or \texttt{reduce} intrinsics to exhibit
parallelism. In returns, it can be efficiently compiled to either CUDA or C++
with calls to the Thrust~\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are dynamically compiled to native code.

@@

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with a Python subset called Pythran to bring fine-grain parallelism to Python
while being backward compatible with both the Python language, and the
sequential algorithm. It means that:
\begin{enumerate}
    \item Any Pythran code can be run (sequentially), with no module dependency or code change,
        by any Python interpreter.
    \item Parallelism is explicit and incrementally added to the original code
        through directives.
\end{enumerate}

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists in a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure.

The languages targeted by OpenMP are statically compiled. This section studies
the semantic adaptation required to use the same API for a scripting language
such as Python.

@@

While Pythran does have a decorator mechanism~\footnote{\see PEP 319,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation fo
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to random instructions.
PLW~\cite{dongara2007} suggests to use string instructions to hold such
decorations. As Python does not have anonymous block~\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.}, one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Alternatively, the syntax
\texttt{if 'my annotation':} is also supported.
Listing~\ref{lst:instruction-annotation} shows how to annotate a parallel loop
with reduction this way.

\begin{figure}
    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
    loop with reduction annotated with OpenMP directive.}]
'omp parallel for reduction(+:r)'
for auto in my_container:
  r += x
  \end{lstlisting}
\end{figure}

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables referred to in a function are to be
considered when building such variable lists: there is no concept of variable
local to a block. Using the \texttt{default(none)} clause is possible to ensure
that no variable gets forgotten when building such lists, as a compile-time
error is raised if a variable is referred in a block but not in the associated
parallel region.

@@

Python also generalizes the concept of iterators through generators, which also
behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
\emph{stop}, \emph{step})}, successively yields value starting from
\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}. It is
easily extended to support random access, but it generally does not make
sense to use a generator as a loop iterator, as the relation between two
random states of the iterator may be of an arbitrary complexity.

A particular class of generator behaves like adaptor, in the sense that they
apply a particular expression on each value of the iterator. It is the case of
the \texttt{enumerate} builtin that yields each element of the enumerated
iterator associated with its index. Generator expressions such as \texttt{(x*x
for x in l)} have the same behavior. These generators are random iterators only
if their input iterator is a random iterator itself.

Finally, if the iterator yields a tuple, it is possible to unpack it inside the
\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
all the unpacked variables are considered as iterators, especially with respect
to default privatization rules. In the given example, \texttt{i} and \texttt{v}
are private, and the parallel iteration is valid because the input of
\texttt{enumerate} is a list, which allows random access iteration.

\begin{figure}
    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
    loop with tuple unpacking.}]
'omp parallel for'
for i, v in enumerate([2, 3, 5, 7, 11]):
  print i, ':', v
  \end{lstlisting}
\end{figure}

@@

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used for portability with system where OpenMP is not
available. Python does not have a preprocessor, but it is possible to catch the
import exception if the \texttt{omp} module is not found.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if():
  import omp
  using = num_threads = 0
  sum = sum2 = 0
  LOOPCOUNT = 1000
  'omp parallel for private(i) if(using == 1)'
  for i in range(LOOPCOUNT + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
  return known_sum == sum and num_threads == 1
    \end{lstlisting}
\end{figure}

@@

Pythran is subset of Python focused on scientific computing. It is implicitly
statically typed and support most Python constructs except those that involve
introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
\texttt{eval}). A few standard module are supported in addition to the core
language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
partially supported. The associated compiler turns Pythran code, eventually
annotated with OpenMP and exported function type annotation, into C++ code. Its
processing is summarized in Figure~\ref{fig:pythran-compiler}. The purpose of
this section is not to dig into the internal of the compiler, but rather to
focus on the impact of the parallelism layer, especially on the runtime library,
\texttt{pythonic}.

@@

\centering
\begin{tikzpicture}[
file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center},
    tool/.style={draw=black!50,fill=black!10,circle, align=center}]
    \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
\node[file] (annotation)	   [right=of python] {\textbf{Type Info}};
\node[file] (omp)	   [left=of python] {\textbf{OpenMP}};
\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
\node[file] (cxx) [below=of pythran] {\textbf{C++}};
\node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
\node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
\node[tool] (gxx) [below=of cxx] {\textbf{g++}};
\node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

@@

A critical point when designing the pythonic runtime was memory management, the
very same aspect that lead to the use of a GIL in CPython. Pythran handles the
problem by refusing recursive types, which makes it possible to solely rely on
reference counting for memory management. It can be implemented through the
thread-safe shared pointer mechanism provided in C++11 standard library.

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. It limits parallelism, so reducing the
number of copies remains an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing may be a trouble. To avoid a reference increment, one can use pass
parameters by reference. Using an interprocedural memory effect analysis, Pythran
determines for each argument whether it must be passed using reference or const
reference to prevent this extra cost.

@@

During Python to C++ translation, Pythran adopts a blind strategy: it does not
understand the semantic of the annotations. Instead, it just split each
annotation into a context-sensitive part ---the variable names--- and a
context-insensitive part ---the clauses---, and attach them to the proper
instruction in the AST. It then ensures a bijective translation between Python
instruction and generated C++ instruction so that the OpenMP directive is
regenerated on the proper instruction. The same approach is used at the
expression level, to be compatible with the \texttt{if} clause.

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, we
successively parallelized it with OpenMP, turned it into Python with a Pythran
compatible kernel and parallelized the Python version using the approach
described in this paper. We also measurer the SLOC for the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It
non only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

@@

Pythran exposes the full OpenMP interface to the user, thus enabling both data
and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
not the case in Cython, where:

\begin{itemize}
    \item Only loops can be made parallel, through the use of a dedicated
        intrinsic \texttt{prange} as loop iterator. To use more advanced
        parallelism construction, one must rely on the possibility to write raw C
        code embedded in cython.~\todo{SG->PB: tu m'explique ce truc de C dans
        cython ?}

    \item Cython automatically infers reductions and variable privacy, giving
        less control to the end-user.

@@

    \item The code included in a parallel region interacts with the Python code
        in a restricted manner. For instance, it is impossible to use a function
        imported from a Python module, while it is still possible to use native
        C functions.
\end{itemize}

Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrates the
difference between the two approaches. It shows that while being more integrated
in Python, the Cython approach is more verbose.

\begin{figure}[ht]

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
    of a parallel reduction.}]
import math
cimport cython
from cython.parallel import parallel, prange
from libc.math cimport sqrt
def sum_sqrt():
    cdef float r = 0
    cdef int i
    with nogil, parallel():
        for i in prange(10000000):
            r += sqrt(i)
    return r
\end{lstlisting}
\end{figure}
%
\begin{figure}[ht]
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython implementation
    of a parallel reduction.}]
#pythran export sum_sqrt()
import math
def sum_sqrt():
    r = 0
    "omp parallel for private(i) reduction(+:r)"
    for i in xrange(10000000):
        r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

The performance of the two approaches is illustrated in
Figure~\ref{fig:cython-pythran}.

@@

The benchmarked codes are typical mathematical, image-processing or linear
algebra kernels. All these kernels have been written in Cython and Python
---compatible with Pythran--- and annotated through the mechanism of each
language to exhibit parallelism, then compiled into native code using GCC 4.1
and the \texttt{-O2 -fopenmp} flag. Their execution time when called from the
Python interpreter is measured through the \texttt{timeit} module. All results
are normalized to Cython sequential execution time.

This figure shows our inability to select proper benchmarking code, and to run
them on a real multiprocessor machine, but this is gonna change!~\todo{SG->PB:
c'est pour toi !}

All the source codes used for these benchmarks are availabe on the Pythran
reporsitory.~\footnote{\emph{cf}.\
\texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
branch.}

@@

This paper studies the relevancy of adding OpenMP annotations to Python code. It
shows that providing minor semantic adaptations, these annotations are relevant
and makes it possible for Python code to benefit from multi-core performance
boost without worrying about the Global Interpreter Lock.

To achieve this goal, a translator from a subset of the Python language to C++
has been designed. This translator can turn regular Python modules
annotated with OpenMP directives and a few type annotations into native parallel
module. Both the input module and the generated one remain compatible with the
standard interpreter.

The approach is compared with Cython, an extension of the Python language used
to generate native module with an hybrid Python-C syntax that also provide means
to exhibits fine grained parallelism. Modules generated by Pythran outperform
modules generated by Cython for various kernels and both benefit from the
parallel boost.

Future works include the extension of the approach to OpenMP 4 and the
\texttt{target} clause that should make it possible to target MIC from Python.
There is also many vectorization possibility in Python, especially in the list
comprehension construction, that needs to be explored.

@@

This work has received fundings from the \texttt{Silkan} company and the French
ANR through the CARP project. The authors would like to thanks Adrien
\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime.
Fabien \textsc{Dagnat} and Mehdi \textsc{Amini} also contributed to this paper
through their many valuable advices.

>>>>

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}

\lstset{language=python,morekeywords={cimport, cdef, with}}

@@

    The Python language offers an interesting solution as a scientific
    computing language thanks to the SciPy package. However, the lack of
    fine-grained parallelism support remains an important shortcoming. This
    paper proposes to add OpenMP directives to the Python language and shows
    that it is possible to turn a module written in a large Python subset
    enhanced with these directives into a native parallel module that runs as
    fast as its naive C equivalent, while keeping the input code totally
    compatible with the python interpreter. The approach is validated through
    the Pythran compiler.

@@

The prohibitive performance overhead of the language implied by its interpreted
and highly dynamic nature does not prevent its usage as a language for high
performance computing. SciPy overcomes this issue through the usage of low-level
routines written in C or Fortran and encapsulated in Python through
\emph{native} modules. Moreover, its core data structure, the multi-dimensional
array~\cite{numpyarray2011}, was designed so that the underlying data are
available to both native modules and the Python interpreter without conversion
cost.

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
the existing approaches and shows a critical lack of parallelism support, and
emphasises on the need for a backward-compatible approach.
Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives
to the Python language in order to add fine-grain parallelism support to Python
in the context of a Python subset called Pythran. Static compilation of this
language into C++ and the underlying runtime is described in
Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
scientific kernels and compared to Cython in Section~\ref{sec:validation}.

@@

In the many cores area, turning to parallelism to balance the performance
limitations of scripting languages, as described 
% in~\cite{choy05} in the context of the MATLAB language, or
in~\cite{mals07} in the context of the R language, seems legitimate. In the
context of Python, most approaches have focused on fork-based parallelism.

@@

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. It's execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. GIL contention actually makes the code run slower.

As a consequence, Python developers need to write multi-threaded native
modules in order to fully benefit from multiple cores. This leads to a kind of
computations referred as hybrid computations.

@@

In the context of interpreted language, \cite{dongara2007} defines a computation
as \emph{hybrid} when part of the code is interpreted, and part of it is
executed natively.

It is now common for scripting language to have C bindings. To take advantage of
compiled code, and to overcome the GIL limitations, Python developers must write
parallel C/C++ functions and the associated boilerplate based on the Python C
API~\cite{pythoncapi}. Tools have been developed to relieve the user from
writing the encapsulating code, notably SWIG~\cite{swig2003} that relies on an
enhanced interface specification, or
\texttt{boost::python}~\cite{boostpython2007} that relies on C++ facilities to
guide translation.

@@

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, among which the possibility to define
parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.
%The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
%compiler uses additional type information to generate sequential LLVM bytecode.
PyCUDA and PyOpenCL~\cite{klockner2012} also mixes python with kernels embedded
as raw strings to target accelerators.

The core issue of these approaches is that they imply to modify, in a more or
less trivial way, the original code. They require to invest in a new language,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code where to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Copperhead~\cite{copperhead2011} is a functional, data parallel language embedded in
Python. It uses n-uplets, numpy arrays and lists as its core data structure and
prohibits usage of many control-flow operators such as loops, enforcing the use
of the \texttt{map}, \texttt{filter} or \texttt{reduce} intrinsics to exhibit
parallelism. In returns, it can be efficiently compiled to either CUDA or C++
with calls to the Thrust~\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.

@@

To be completely backward compatible, it must be possible to run the input code
in an environment that is not aware of the existence of the parallelizing
solution. This principle is not respected by parallel libraries, but code
annotations satisfy it. This is where  OpenMP~\cite{openmp3.1} parallel
annotations shines: the original code remains compatible with a compiler that is
not aware of OpenMP.

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with a Python subset called Pythran to bring fine-grain parallelism to Python.
while being backward compatible with both the Python language, and the
sequential algorithm.It means that:
%
\begin{enumerate}
    \item Any Pythran code can be run (sequentially), with no module dependency or code change,
        by any Python interpreter.
    \item Parallelism is explicit and incrementally added to the original code
        through directives.
\end{enumerate}

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists in a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure. The languages targeted by OpenMP are statically
compiled. This section studies the semantic adaptation required to use the same
API for a scripting language such as Python.

@@

While Pythran does have a decorator mechanism~\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to random instructions.
PLW~\cite{dongara2007} suggests to use string instructions to hold such
decorations. As Python does not have anonymous block,
%~\footnote{\see PEP340, \url{http://www.python.org/dev/peps/pep-0340/} for a
%discussion concerning anonymous block support in Python.}
one has to create a dummy \texttt{if 1:} instruction to apply an annotation to a
whole block. Alternatively, the syntax \texttt{if 'my annotation':} is also
supported.

%Listing~\ref{lst:instruction-annotation} shows how to annotate a
%parallel loop with reduction this way.
%
%\begin{figure}
%    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
%    loop with reduction annotated with OpenMP directive.}]
%'omp parallel for reduction(+:r)'
%for auto in my_container: r += x
%  \end{lstlisting}
%\end{figure}

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables referred to in a function are to be
considered when building such variable lists: there is no concept of variable
local to a block. Using the \texttt{default(none)} clause is possible to ensure
no variable gets forgotten when building such lists.

@@

Python also generalizes the concept of iterators through generators, which 
behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
\emph{stop}, \emph{step})}, successively yields value starting from
\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}. It is
easily extended to support random access, but it generally does not make
sense to use a generator as a loop iterator, as the relation between two
random states of the iterator may be of an arbitrary complexity.

A particular class of generator behaves like adaptor, in the sense that they
apply a particular expression on each value of the iterator. It is the case of
the \texttt{enumerate} builtin, that yields each element of the enumerated
iterator associated with its index, and of generator expressions such as
\texttt{(x*x for x in l)}. These generators are random iterators only if their
input iterator is a random iterator itself.

Finally, if the iterator yields a tuple, it is possible to unpack it inside the
\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
all the unpacked variables are considered as iterators, especially with respect
to default privatization rules: in the given example, \texttt{i} and \texttt{v}
are private, and the parallel iteration is valid because the input of
\texttt{enumerate} is a list, which allows random access iteration.

\begin{figure}
    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
    loop in Pythran with tuple unpacking.}]
'omp parallel for'
for i, v in enumerate([2, 3, 5, 7, 11]): print i, ':', v
  \end{lstlisting}
\end{figure}

@@

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used to detect when OpenMP is not available. Python does
not have a preprocessor, but it is possible to catch the import exception if the
\texttt{omp} module is not found.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if(loop_count):
  import omp
  using = sum = sum2 = 0
  'omp parallel for if(using == 1)'
  for i in range(loop_count + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (loop_count * (loop_count + 1)) / 2
  return known_sum == sum and num_threads == 1
    \end{lstlisting}
\end{figure}

@@

Pythran is subset of Python focused on scientific computing. It is implicitly
statically typed and support most Python constructs except those that involve
introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
\texttt{eval}). A few standard module are supported in addition to the core
language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
partially supported. The associated compiler turns Pythran code, eventually
annotated with OpenMP and exported function type annotation, into C++ code. Its
processing is summarized in Figure~\ref{fig:pythran-compiler} and relies on type
inference to remove most dynamic behavior, combined with a high-level runtime
library, \texttt{pythonic}, to allow a one-to-one mapping between Python and C++
constructs. The purpose of this section is not to dig into the internal of the
compiler, but rather to focus on the impact of the parallelism layer, especially
on the runtime library.

@@

\centering
\begin{tikzpicture}[
file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
node distance=0.6cm},
    tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
distance=0.6cm}]
\node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
\node[file] (annotation)	   [right=of python] {\textbf{Type Info}};
\node[file] (omp)	   [left=of python] {\textbf{OpenMP}};
\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
\node[file] (cxx) [below=of pythran] {\textbf{C++}};
\node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
\node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
\node[tool] (gxx) [below=of cxx] {\textbf{g++}};
\node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

@@

A critical point of the design of the \texttt{pythonic} runtime is memory
management, the very same aspect that lead to the use of a GIL in CPython.
Memory management is implemented in Shedskin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
object inside parallel regions, thus making memory management explicit for these
parts.

Pythran handles the problem by refusing recursive types, which makes it possible
to solely rely on reference counting for memory management. It can be
implemented through the thread-safe shared pointer mechanism provided in C++11
standard library.

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still imply copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this extra cost.

@@

During Python to C++ translation, Pythran adopts a blind strategy: it does not
understand the semantic of the annotations. Instead, it just split each
annotation into a context-sensitive part ---the variable names--- and a
context-insensitive part ---the clauses---, and attach them to the proper
instruction in the AST.

The Pythran compiler ensures a bijective translation between Python instruction
and generated C++ instruction so that the OpenMP directive is regenerated on the
proper instruction. The same approach is used at the expression level, to be
compatible with the \texttt{if} clause.

At the AST level, it means that the Python AST is first reduced to a Python AST
where all nodes not available in C++ are transformed into a compatible
representation. For instance, list comprehension expressions are turned into
function calls or tuple unpacking is turned in multiple assignments. During this
transformation process, an expression is always transformed into a single expression
and a statement is always transformed into a single statement. Then this
AST is converted into a C++ AST meant to be pretty-printed.

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, Hyantes,
we successively parallelized it with OpenMP, turned it into Python with a
Pythran compatible kernel and parallelized the Python version using the approach
described in this paper. We also measurer the SLOC for the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It non
only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

The Hyantes programs scales relatively well. Figure~\ref{fig:hyantes-speedup} shows
the relative speedup of the application when increasing the number of OpenMP
threads.

\begin{figure}[ht]
    \caption{Relative speedups of Pythran+OpenMP version of the Hyantes prototype.}
    \label{fig:hyantes-speedup}
    \centering
    \includegraphics[width=\textwidth]{hyantes}
\end{figure}

@@

Pythran exposes the full OpenMP interface to the user, thus enabling both data
and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
not the case in Cython, where:
%
\begin{itemize}
    \item Only loops can be made parallel, through the use of a dedicated
        intrinsic \texttt{prange} as loop iterator.

    \item Reductions and variable privacy are inferred, but it chokes on
        reduction on private variables.

@@

    \item It is impossible to use a function imported from a Python module in a
        parallel region, while it is still possible to use native C functions.

\end{itemize}

Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrate the
difference between the two and illustrates the intrusive behavior of Cython.

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
    of a parallel reduction.}]
from libc.math cimport sqrt
from cython.parallel import parallel, prange
def sum_sqrt(double r):
    cdef int i
    with nogil, parallel():
        for i in prange(10000000): r += sqrt(i)
    return r
    \end{lstlisting}
%
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython implementation
    of a parallel reduction.}]
#pythran export sum_sqrt(float)
import math
def sum_sqrt(r):
    "omp parallel for private(i) reduction(+:r)"
    for i in xrange(10000000): r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

The performance of the two approaches is shown in
Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
mathematical, image-processing or linear algebra kernels. All these kernels have
been written in Cython and Python ---compatible with Pythran--- and annotated
through the mechanism of each language to exhibit parallelism, then compiled
into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag. Their
execution time when called from the Python interpreter is measured through the
\texttt{timeit} module on a laptop with 4 Intel i7 cores at 2.10 GHz with
hyperthreading deactivated. All results are normalized against Cython sequential
execution time. They show that while being of a higher level than Cython,
Pythran achieves comparable results.

The source codes used for these benchmarks are available on the Pythran
repository.~\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
branch.}

@@

@@

This paper studies the relevancy of adding OpenMP annotations to Python code. It
shows that providing minor semantic adaptations, these annotations make it
possible for Python code to benefit from multi-core performance boost while
retaining backward compatibility and without worrying about the Global
Interpreter Lock.

To achieve this goal, a translator from a subset of the Python language to C++
has been designed. This translator turns regular Python modules annotated with
OpenMP directives and a few type annotations into native parallel module. The
input module remains compatible with the standard interpreter.

The approach is compared with Cython, an extension of the Python language used
to generate native module with an hybrid Python-C syntax that also provide means
to exhibit fine grained parallelism. It shows that retaining Python
compatibility does not prevent the achievement of comparable performances.

Future work includes the extension of the approach to OpenMP 4 and the
\texttt{target} clause that should make it possible to target MIC from Python.
There is also many vectorization possibility in Python, especially in the list
comprehension construction, that needs to be explored.

@@

This work has received fundings from the \texttt{Silkan} company and the French
ANR through the CARP project. The authors would like to thanks Adrien
\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime,
Albert \textsc{Cohen}, B{\'e}atrice \textsc{Creusillet}, Fabien
\textsc{Dagnat} and Mehdi \textsc{Amini} for their valuable advices.
